Live coding environment for a data exploration language

### 2019-02-10 wip-live-programming/paper.tex
Related work

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
can even export the workflow as a script \cite{wrangler}, but data anlysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users have to structure code in
cells and manually reevaluate cells after a change. Those systems also do not track
dependencies between cells, which can lead to inconsistencies; an issue that has been addressed
by e.g.~\citet{dataflow} or \citet{wrattler}.

@@

\begin{figure}
\begin{equation*}
\hspace{-3.5em}
\begin{array}{l}
\inference[\rname{mem}~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
  \bndclr{v_0} \vdash \{.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..\} & \bndclr{v_i} \vdash \tau_i }
  { \bnd{mem}(m, s) \vdash \tau }
\\[1.5em]
\inference[\rname{var}~]
  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
  { \bnd{var}(x, s) \vdash \tau' }
\\[1.5em]
\inference[\rname{fun}~]
  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E \\
    \bndclr{v_c} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} &
    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
\end{array}
\end{equation*}
\caption{Rules that define type checking of terms and expressions over a dependency graph $(V, E)$}
\label{fig:tc}
\end{figure}

@@

The sample data analysis by Financial Times in Section~\ref{sec:background-jupyter} illustrates why
notebook users often avoid abstraction. Wrapping code in a function would make it impossible to
split code into cells, and see results of intermediate steps. Instead, the sample analysis used a
global variable at the top of the notebook with possible values in a comment.

@@

\noindent
As before, a program is a sequence of commands and a command can be either a let binding or a term.
Additionally, a command can be a labelled sequence of commands or a reference to a label. When
evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of commands $p$
before any other reduction occurs. Consequently, variables used in the labelled block are
dynamically scoped and we can use let binding to redefine a value of a variable before invoking
the block repeatedly. The correct use of dynamically scoped variables can be checked using
coeffects \cite{coeffects}.

This minimalistic abstraction mechanism supports code reuse without affecting how live previews
are computed. The commands in the labelled block still require all let bound variables to be
defined before the block. Those define sample data for development and can be later redefined before
reusing the block. This functionality is not yet available in the data exploration environment
discussed in Section~\ref{sec:evaluation-case}, but we intend to add it in a future version.

@@

This paper approaches the problem of live coding environments from a theoretical programming
language perspective with a special focus on tooling for data science. Hence, the related and
future work spans numerous areas.

\subsubsection*{Design and human-computer interaction.}
TODO: \citet{principle}

From a design perspective, the idea of live programming environments has been popularised by
Bret Victor \cite{learnable}. Active research on novel forms of interaction happens in areas
such as live coded music \cite{beyond,sonic}.
The idea of live previews can be extended to direct manipulation \cite{direct}. The Gamma
provides limited support for directly manipulating data (Section~\ref{sec:live}), but we
intend to explore this direction further.

\paragraph{Data science tooling.}
An essential tool in data science is REPL (read-eval-print-loop) \cite{drscheme}, which is now
widely available. This has been integrated with rich graphical outputs in tools such as Jupyter
notebooks \cite{jupyter,ipython}, but such previews are updated using an explicit command.
Integrating our work with Jupyter to provide instant live previews for R or Python would be
an interesting extension of the presented work.

\paragraph{Live coding and live previews.}
Live previews have been implemented in LightTable \cite{lighttable} and, more recently, in
editors such as Chrome Developer Tools, but neither presents a simple description of their
inner workings. An issue that attracts much attention is keeping state during code edits
\cite{alive,livingit}. This would be an interesting problem if we extended our work to
event-based reactive programming.

\paragraph{Structured editing.}
An alternative approach to ours is to avoid using text editors. Structured editors
\cite{structure-based} allow the user to edit the AST and could, in principle, recompute previews
based on the performed operations, or preview evaluation as in interactive functional programming
\cite{interactive}. A promising direction is using bi-directional lambda calculus~\cite{hazelnut}.
Finally, abandoning text also enables building richer, more human-centric
abstractions as illustrated by Subtext~\cite{subtext}. Our current focus, however, remains
on text-based editors.

\paragraph{Dependency analysis.}
Our use of dependency graphs \cite{dependencies} is first-order. Building dependency graphs
involving function calls using modern compiler methods \cite{optimizing} or program slicing
\cite{slicing} would allow us to deduce possible inputs for functions and use those for
previews rather than changing the language as suggested in Section~\ref{sec:extra-abstraction}. This direction
is worth considering, but it requires more empirical usability testing.

\paragraph{Semantics and partial evaluation.}
The evaluation of previews can be seen as a form of partial evaluation \cite{partial}, done in a
way that allows reuse of results. This can be done implicitly or explicitly in the form of
multi-stage programming \cite{metaml}. Both can provide useful perspective for formally analysing
how previews are evaluated. Semantically, the evaluation of previews can be seen as a modality
\cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt}, which,
in turn, can be understood in terms of comonads \cite{cmtt-denotation}. This provides an intriguing
direction for rigorous analysis of the presented system.

>>>>

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
can even export the workflow as a script \cite{wrangler}, but data anlysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users structure code in
cells and manually reevaluate cells. Many notebook systems do not track dependencies
between cells, which can lead to inconsistencies; an issue that has been addressed
by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.

@@

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\hspace{-3.5em}
\begin{array}{l}
\inference[\rname{mem}~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
  \bndclr{v_0} \vdash \{.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..\} & \bndclr{v_i} \vdash \tau_i }
  { \bnd{mem}(m, s) \vdash \tau }
\\[1.5em]
\inference[\rname{var}~]
  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
    \bndclr{v} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} & \tau_i=\tau'\rightarrow\tau'' }
  { \bnd{var}(x, s) \vdash \tau' }
\\[1.5em]
\inference[\rname{fun}~]
  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E \\
    \bndclr{v_c} \vdash \{.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..\} &
    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Rules that define type checking of terms and expressions over a dependency graph $(V, E)$}
\label{fig:tc}
\vspace{-0.5em}
\end{figure}

@@

The data analysis by Financial Times in Section~\ref{sec:background-jupyter} illustrates why
notebook users often avoid abstraction. Wrapping code in a function makes it impossible to
split code into cells and see results of intermediate steps. Instead, the analysis used a
global variable with possible values in a comment.

@@

\noindent
We introduce two new kinds of commands: a labelled sequence of commands and a referenceto a label.
When evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of
commands $p$ before any other reductions. Consequently, variables used in the labelled block are
dynamically scoped and we can use let binding to redefine a value of a variable before invoking
the block repeatedly. The correct use of dynamic scoping can be checked using
coeffects \cite{coeffects}.

This minimalistic abstraction mechanism supports code reuse without affecting how live previews
are computed. Commands in a labelled block require variables to be defined before the block.
Those define sample data for development and can be redefined before reusing the block. We
intend to implement this mechanism in a future version of our data
exploration environment (Section~\ref{sec:evaluation-case}).

@@

We aim to make creating transparent data analyses, such as those done by journalists, easier and we
present a tool that provides instantaneous feedback when writing code in a text editor. Many our
design decisions have alternatives that have been explored in a rich body of literature.

\subsubsection*{Notebooks and data science tools.}
Basic feedback mechanism, the REPL (read-eval-print-loop), originated in Lisp
\cite{lisp} and is now common in many languages \cite{drscheme}. Notebook systems
such as Jupyter and R markdown \cite{rmarkdown,jupyter,ipython} structure code in cells,
but still rely on explicitly executing individual cells using a stateful REPL environment.
Integrating provenance tracking with notebooks \cite{dataflow,noworkflow,wrattler} addresses
the reproducibility issues of this model, but keeps cells as units of work. Ideas such as
dependency tracking and efficient recomputation exist in many visual data exploration tools
\cite{control,tableau,vizdom} and scientific workflow systems \cite{taverna,kepler}. Some of the
implementation techniques are related to our work, but we focus on text-based scripts.
Tempe \cite{tempe} focuses on streaming data, but provides a text-based scripting environment
with automatic code update and \citet{ripple} empirically evaluate its usability in contrast to
REPLs. The challenges faced by Tempe are more akin to dynamic software update \cite{dsu}.

\subsubsection*{Live coding and editing.}
Live coding based on textual programs has been popularised by \citet{principle,learnable} and
is actively developed in domains such as live coded music \cite{beyond,sonic}. \citet{liveroad}
reviews the use of live coding in a Smalltalk derived environment. Lighttable by \citet{lighttable}
and Chrome DevTools provide limited live previews akin to those presented in this paper,
but without well specified recomputation model. Finally, work on keeping state during
code edits \cite{alive,livingit} would be relevant for supporting streaming data. In the future,
we also plan to explore how direct manipulation \cite{direct} could be integrated with our preview
mechanism. \citet{liveroad}

Using structured editors \cite{structure-based} rather than text editors is a promising alternative
direction as editors can recompute program based on understanding of edit operations
\cite{interactive,livenut,lamdu}. This can be elegantly implemented using bi-directional
lambda calculus \cite{hazelnut}, but it also makes us consider more human-centric abstractions
\cite{subtext,directprog} along the lines of the one outlined in Section~\ref{sec:extra-abstraction}.

\subsubsection*{Incremental computation and dependency analysis.}
Work on self-adjusting and incremental computation \cite{selfadjusting,adapton} handles
recomputation when the program stays the same, but input changes. It relies on dependency tracking
similar to ours. We rely on static code analysis whereas most incremental systems,
e.g.~\cite{incremental,adaptivefp,deltaml} evaluate the program and use programmer-supplied
information to build a dependency graph. \citet{adapton} implement a small adaptive interpreter
that treats code as changing input data, suggesting an interesting implementation technique for
live coding systems. Our use of dependency graphs \cite{dependencies} is static and first-order
and can be seen as a form of program slicing \cite{slicing}, although our binding process is
more directly inspired by Roslyn \cite{roslyn}, which uses it for efficient background type-checking.

\subsubsection*{Semantics and partial evaluation.}
The evaluation of previews can be seen as a form of partial evaluation \cite{partial}, done in a
way that allows reuse of results. This can be done implicitly or explicitly in the form of
multi-stage programming \cite{metaml}. Both can provide useful perspective for formally analysing
how previews are evaluated. Semantically, the evaluation of previews can be seen as a modality
\cite{modal} and delayed previews are linked to contextual modal type theory \cite{cmtt}, which,
in turn, can be understood in terms of comonads \cite{cmtt-denotation}. This provides an intriguing
direction for rigorous analysis of the presented system.
